---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This vignette will show the basic use-cases for `RGenEDA` on bulk RNASeq data, however, the package can be easily used for other types of omics data as long as a counts matrix and metadata are present. 

We will rely on data in the `pasilla` dataset and follow the [deseq2 framework](https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

```{r setup}
# Load libraries
library(RGenEDA)
library(DESeq2)
library(pasilla)
```

```{r read in the data}
# Set path to pasilla data
datafile <-  system.file( "extdata/pasilla_gene_counts.tsv", package="pasilla" )

# Load counts
count.table <-  read.table( datafile, header=TRUE, row.names=1, quote="", comment.char="" )
head(count.table)

```

```{r set metadata}
cond.type <-  c( "untreated", "untreated", "untreated","untreated", "treated", "treated", "treated" )
lib.type   <-  c( "single-end", "single-end", "paired-end", "paired-end", "single-end", "paired-end", "paired-end" ) 

metadata <- data.frame(condition = cond.type,
                       library = lib.type)
rownames(metadata) <- colnames(count.table)
```


```{r prepare for DESeq2}

# Create a DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = count.table,
                              colData = metadata,
                              design = ~condition + library + condition:library)

# Set reference levels
dds$condition <- relevel(dds$condition, ref = "untreated")
dds$library <- relevel(dds$library, ref = "single-end")

# Pre filter
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

# Run DESeq2
dds <- DESeq(dds)

# Variance stabilizing transform
vsd <- vst(dds)

# Extract matrix
mat <- assay(vsd)
```

Now that we successfully have a matrix and metadata, we are ready to start exploratory data analysis. 
Let's start by looking at the sample distances to see how our samples are clustering. For `RGenEDA` it is useful to set a figure directory where all EDA plots will live. We want to visualize all possible metadata for this. 

```{r sample distances}

# Set output directory
outputDir <- c("/users/mike/Desktop/RGenEDA/img/")

# Set colors for our metadata categories
metaColors <- list(condition = c("untreated" = "blue",
                                 "treated" = "red"),
                   library = c("single-end" = "maroon",
                               "paired-end" = "gold"))

# Call the distance heatmap function
RGenEDA::distanceHeatmap(MAT = mat,
                META = metadata,
                FEATURES = c("condition", "library"),
                PALETTES = metaColors,
                OUTPUT = outputDir)



```

```{r run PCA functions}

# Explore variance
variance <- plotVariance(MAT = mat,
                         OUTPUT = outputDir,
                         LOG = FALSE,
                         nFeaturesDrop = NULL)

# Generate PCs
pcaRes <- generatePCs(MAT = mat,
                      VARS = variance,
                      NFEATURES = 2000)

# Extract PC data
pcaDF <- as.data.frame(pcaRes[["Loadings"]])

# Append metadata to pcaDF
pcaDF <- cbind(pcaDF, metadata)

# Plot PCA
```


```{r eigencorrelations}

# Use Eigencorrelations
ecs <- eigencorr(MAT = mat,
                 META = metadata,
                 NUM_PCS = 4,
                 OUTPUT = outputDir)




```

```{r MA plot}
# Extract DESeq2 results
resultsNames(dds)
results <- as.data.frame(results(dds, name = "condition_treated_vs_untreated"))

MAplot(results = results,
       numerator = "treated",
       refLevel = "untreated",
       log2FC_thresh = 1,
       padj_thresh = 0.05,
       title = "MA_demo",
       figDir = outputDir,
       "MAplotDemo.png")

```
